---
title: "DEGgo"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{DEGgo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

<div align="center">

<img src="../man/figures/DEGgo_logo.svg" width="500" alt="DEGgo logo"/>

<br><br>

<strong>
An integrated R framework for automated bulk RNA-seq differential expression,
functional enrichment, circadian rhythmicity analysis, and reproducible reporting.
</strong>

</div>

## Overview

**DEGgo** is an open-source R package for reproducible bulk RNA-seq analysis.

It integrates:

- input validation and sample matching;
- raw and post-filtering quality control;
- differential expression analysis with DESeq2, edgeR, or limma;
- single and pairwise contrasts;
- PCA, volcano plots, heatmaps, and gene-expression visualization;
- Gene Ontology enrichment;
- circadian rhythmicity analysis using MetaCycle and cosinor regression;
- automated HTML, PDF, and PowerPoint reporting;
- reproducibility files, analysis manifests, and session information.

## Installation

```{r installation}
install.packages("remotes", repos = "https://cloud.r-project.org")
remotes::install_github("ymbouamboua/DEGgo")
```

```{r load-package}
library(DEGgo)
```

Optional rhythmicity dependencies:

```{r install-rhythmicity-dependencies}
install.packages(c("MetaCycle", "cosinor", "cosinor2"))
```

## Workflow

```text
Raw counts
      │
      ▼
 Input validation
      │
      ▼
 Sample QC
      │
      ▼
 Expression filtering
      │
      ▼
 Differential expression
      │
      ├── PCA
      ├── Volcano plots
      ├── Heatmaps
      ├── GO enrichment
      └── Reports
      │
      ▼
 Optional rhythmicity analysis
      ├── MetaCycle
      ├── Cosinor
      ├── Differential rhythmicity
      └── Publication figures
```

## Input data

### Count table

DEGgo accepts raw count tables or matrices. The count table should contain one gene identifier column and one column per sample.

```text
gene_id          gene_name    Sample1    Sample2    Sample3
ENSG00000000003  TSPAN6       120        145        98
ENSG00000000005  TNMD         65         80         50
ENSG00000000419  DPM1         12         18         250
```

### Metadata

Metadata must contain one row per sample. The sample identifier column is specified with `sample_col`.

```text
sample      condition    batch
Sample1     control      A
Sample2     treated      A
Sample3     control      B
```

Sample identifiers in the metadata must match the sample columns in the count table.

# Differential expression workflow

## Example data

DEGgo includes a ready-to-use RNA-seq dataset derived from the Bioconductor [airway](https://bioconductor.org/packages/airway/) package.

```{r airway-data}
counts <- read.delim(
  system.file("extdata", "airway_counts.tsv", package = "DEGgo"),
  check.names = FALSE
)

metadata <- read.delim(
  system.file("extdata", "airway_metadata.tsv", package = "DEGgo"),
  check.names = FALSE
)
```

## Single comparison mode

```{r de-single-mode}
metadata$condition <- metadata$dex

results <- run_deggo(
  counts = counts,
  metadata = metadata,
  project_name = "Airway dexamethasone analysis",
  gene_col = "gene_id",
  organism = "human",
  sample_col = "SampleName",
  method = "DESeq2",
  analysis_mode = "single",
  design_formula = ~ cell + condition,
  contrast = c("condition", "trt", "untrt"),
  filter_method = "count",
  min_count = 5,
  min_samples = 2,
  min_total = 10,
  padj_cutoff = 0.05,
  logfc_cutoff = 0.25,
  output_dir = "DEGgo_airway",
  generate_report = TRUE,
  report_formats = "html",
  generate_pptx = TRUE
)
```

Positive `log2FoldChange` values represent higher expression in the treated group.

## Pairwise comparison mode

```{r pairwise-metadata}
metadata_mouse$group <- interaction(
  metadata_mouse$treatment,
  metadata_mouse$sex,
  metadata_mouse$tissue,
  sep = "_",
  drop = TRUE
)
```

```{r pairwise-contrasts}
pairwise_contrasts <- list(
  treatment_vs_control_male = c(
    "group",
    "treatment_male",
    "control_male"
  ),

  treatment_vs_control_female = c(
    "group",
    "treatment_female",
    "control_female"
  ),

  male_vs_female_control = c(
    "group",
    "control_male",
    "control_female"
  )
)
```

```{r de-pairwise-mode}
pairwise_results <- run_deggo(
  counts = counts_mouse,
  metadata = metadata_mouse,
  project_name = "Pairwise analysis",
  organism = "mouse",
  method = "DESeq2",
  analysis_mode = "pairwise",
  sample_col = "sample",
  design_formula = ~ group,
  pairwise_group_cols = c("treatment", "sex", "tissue"),
  pairwise_contrast_col = "group",
  pairwise_contrasts = pairwise_contrasts,
  output_dir = "DEGgo_pairwise_results",
  generate_report = TRUE,
  report_formats = "html",
  generate_pptx = TRUE
)
```

# Quality control

```{r raw-qc}
qc <- explore_bulk_rnaseq(
  counts = counts,
  metadata = metadata,
  sample_col = "SampleName",
  markers = c("TSPAN6", "DPM1", "SCYL3"),
  output_dir = "DEGgo_airway_QC_raw"
)
```

```{r remove-flagged-samples}
cleaned <- remove_flagged_samples(
  counts = counts,
  metadata = metadata,
  qc_table = qc$qc,
  sample_col = "SampleName",
  remove_col = "recommend_remove",
  gene_cols = "gene_id",
  verbose = TRUE
)
```

# Gene expression extraction

```{r extract-expression}
expr_df <- extract_expression(
  dds = results$dds,
  metadata = results$metadata,
  genes = c("TSPAN6", "DPM1", "SCYL3"),
  assay = "vst",
  gene_col = "SYMBOL"
)
```

```{r plot-one-gene}
plot_gene_expression(
  expr_df,
  gene = "TSPAN6",
  x = "condition",
  color = "condition",
  facet = "cell",
  geom = "violin"
)
```

# Gene Ontology enrichment

```{r run-go-enrichment}
go <- run_go_enrichment(
  sig_deg = results$sig_deg,
  comparison = "condition_trt_vs_untrt",
  ontology = "BP",
  orgdb = org.Hs.eg.db::org.Hs.eg.db,
  output_dir = file.path("DEGgo_airway", "GO_condition_trt_vs_untrt")
)
```

```{r plot-go-terms}
plot_go_terms(
  go_df = go$go_results,
  comparison = "Dexamethasone treated vs untreated",
  top_n = 15
)
```

# Circadian rhythmicity analysis

## Public circadian example

DEGgo includes a reproducible workflow based on GSE98965 from Mure *et al.* (Science, 2018).

```{r public-circadian-example}
public_rhythm <- run_public_circadian_example(
  tissue_code = "LIV",
  output_dir = "Circadian_results"
)
```

```{r prepare-public-circadian-data}
gse98965 <- download_gse98965(
  output_dir = "data/GSE98965"
)

liver_data <- prepare_circadian_tissue(
  expression_data = gse98965,
  tissue_code = "LIV"
)
```

## Run rhythmicity analysis

```{r run-rhythmicity}
rhythm_results <- run_deggo_rhythmicity(
  expr = expression_matrix,
  metadata = metadata,
  sample_col = "sample",
  time_col = "time",
  assay = "log2_normalized",
  methods = c("meta2d", "cosinor"),
  period_range = c(20, 28),
  cycle_length = 24,
  padj_cutoff = 0.05,
  output_dir = "DEGgo_rhythmicity_results"
)
```

## Extract rhythmic genes

```{r extract-rhythmic-genes}
rhythmic_genes <- deggo_extract_rhythmic_genes(
  rhythm_results,
  rhythmic_by = "both",
  padj_cutoff = 0.05
)
```

## Differential rhythmicity

```{r differential-rhythmicity}
group_rhythm <- run_deggo_rhythmicity(
  expr = expression_matrix,
  metadata = metadata,
  sample_col = "sample",
  time_col = "time",
  group_col = "condition",
  assay = "log2_normalized",
  methods = c("meta2d", "cosinor"),
  cycle_length = 24,
  padj_cutoff = 0.05,
  output_dir = "DEGgo_group_rhythmicity"
)
```

# Main functions

| Function | Description |
|:---------|:------------|
| `run_deggo()` | Run the complete differential expression workflow |
| `run_deggo_rhythmicity()` | Detect rhythmic genes with MetaCycle and cosinor regression |
| `deggo_extract_rhythmic_genes()` | Extract rhythmic genes from rhythmicity results |
| `download_gse98965()` | Download the public GSE98965 circadian dataset |
| `prepare_circadian_tissue()` | Prepare one tissue from the public circadian atlas |
| `run_public_circadian_example()` | Run the complete public circadian example |
| `check_raw_counts()` | Validate counts and sample identifiers |
| `explore_bulk_rnaseq()` | Perform exploratory quality control |
| `remove_flagged_samples()` | Remove low-quality samples |
| `run_sample_qc()` | Perform post-filtering sample quality control |
| `run_go_enrichment()` | Perform Gene Ontology enrichment |
| `plot_heatmap()` | Plot differentially expressed genes |
| `plot_pca()` | Generate a PCA plot |
| `plot_volcano()` | Generate a volcano plot |
| `generate_deggo_report()` | Generate HTML or PDF reports |
| `generate_deggo_pptx()` | Generate a PowerPoint summary |
| `deggo_palette()` | Retrieve DEGgo colour palettes |

# Supported organisms

| Organism | Parameter | OrgDb package |
|:---------|:----------|:--------------|
| Human (*Homo sapiens*) | `"human"` | `org.Hs.eg.db` |
| Mouse (*Mus musculus*) | `"mouse"` | `org.Mm.eg.db` |
| Rat (*Rattus norvegicus*) | `"rat"` | `org.Rn.eg.db` |

Additional organisms can be analyzed with a compatible Bioconductor `OrgDb` object.

# Citation

If you use DEGgo, please cite the Zenodo release:

<https://doi.org/10.5281/zenodo.20785178>

# Contributing

Bug reports, feature requests, and pull requests are welcome through the [DEGgo GitHub repository](https://github.com/ymbouamboua/DEGgo).

# License

MIT License © Yvon MBOUAMBOUA
